---
title: "RMS Titanic"
author: "Sparsh Gupta"
date: 2025-01-30
output:
  github_document:
    toc: true
prerequisites:
  - e-vis01-bar-charts
---

*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|-------------------|----------------------------|-------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

```{r setup}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look

<!-- -------------------------------------------------- -->

### **q1** Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
## TASK: Perform a `glimpse` of df_titanic
glimpse(df_titanic)
```

**Observations**:

-   Class: This represents the class the passengers were traveling in.
-   Sex: This represents the sex of the individuals. ("Male" or "Female").
-   Age: This represents the age of the observations.
-   Survived: This represents if the observation survived the accident or not. ("Yes" if survived, otherwise "No").
-   n: This number represents the number of individuals in that specific observation category.

### **q2** Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of souls aboard. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

**Observations**:

-   From the R code above, the total number of souls aboard the RMS Titanic is 2201. This value represents the sum of all recorded individuals in the dataset.
-   According to Wikipedia, the total number of people aboard the Titanic was 2224.
-   Are there any differences?
    -   There is a difference of 23 individuals between the calculated dataset total (2201) and the Wikipedia source (2224).
    -   The difference (23 individuals) represents approximately 1.03% of the total, which is relatively small when compared to the number of souls aboard.
-   If yes, what might account for those differences?
    -   The dataset used for the computation might not account for every individual aboard. One possible things that could cause this are crew members or passengers that boarded the Titanic after the ticket records were finalized. Another possible cause might be unofficial passengers aboard who were not registered.
    -   Titanic's records were created over 100 years ago, and therefore, slight inconsistencies between data records are not uncommon.
    -   There is a possibility that certain rows or individuals may have been omitted during preprocessing of data, for reasons such as incomplete data for some observations.
    -   There might be errors in defining who was considered "aboard" because it might not account for crew members whose roles were clearly less documented (e.g. temporary hires, etc.)
    -   The Titanic also made some port stops on the trip before the accident happened, where it picked up passengers and dropped some passengers off. This means that the number of passengers aboard the Titanic fluctuated and so maintaining a definite count of people aboard might not have been possible.

### **q3** Create a plot showing the count of persons who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}
## TASK: Visualize counts against `Class` and `Sex`
df_titanic %>%
  filter(Survived == "Yes") %>%
  ggplot(aes(Class, n, fill = Sex)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Count of Survivors by Class and Sex",
    x = "Class",
    y = "Number of Survivors",
    fill = "Sex"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    strip.text = element_text(size = 12, face = "bold")
  )

```

**Observations**:

-   The bar plot represents the count of survivors grouped by "Class" (1st, 2nd, 3rd, Crew) and divided by "Sex" (Male, Female). Each "Class" has separate bars for males and females to compare survival counts.
-   Class-wise Observations:
    -   1st Class: Female survivors significantly outnumber male survivors. This suggests strong prioritization of women in lifeboats for this class, possibly due to social norms or accessibility advantages.
    -   2nd Class: Female survivors are higher than male survivors, but the difference is not as large as in 1st class. This indicates that while women were prioritized, the survival advantage for females was lower than in 1st class.
    -   3rd Class: Male and female survivors appear almost equal in number. This contrasts with the common expectation that females had a higher survival rate, suggesting barriers to lifeboat access for women in 3rd class.
    -   Crew: Male survivors overwhelmingly outnumber female survivors. This likely reflects the fact that the Titanic crew had significantly more men than women (to explore this further, we need to look at the number of males vs females in the crew). Another possible reason is that male crew members may have had better access to lifeboats.
-   Females had higher survival rates in 1st and 2nd class, which is probably because of social norms in higher classes where saving women and children first was the priority.
-   In 3rd class, survival rates between males and females are more balanced, possibly due to difficulties accessing lifeboats.
-   Overall, the data suggests a class-based and gender-based disparity in survival, with higher-class passengers having better survival chances.

# Deeper Look

<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop
```

### **q4** Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3. Is there anything *fishy* in your plot?

```{r q4-task}
df_prop %>%
  filter(Survived == "Yes") %>%
  ggplot(aes(Class, Prop, fill = Sex)) +
  geom_bar(stat = "identity", position = "dodge", na.rm = TRUE) +
  labs(
    title = "Proportion of Survivors by Class and Sex",
    x = "Class",
    y = "Survival Proportion",
    fill = "Sex"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    strip.text = element_text(size = 12, face = "bold")
  )

```

**Observations**:

-   The bar plot represents the proportion of passengers categorized by Class (1st, 2nd, 3rd, Crew) and divided by Sex (Male, Female).
-   Class-wise Observations:
    -   1st Class: The survival proportion is 1.0 for both males and females, meaning all recorded survivors from this class survived at the same rate. This differs from the q3 plot, where the absolute count of surviving females was higher.
    -   2nd Class: Again, both males and females have a survival proportion of 1.0, meaning all survivors from this class, regardless of sex, had the same survival rate.
    -   3rd Class: Females have a higher survival proportion than males, meaning a larger fraction of female passengers in this class survived compared to males. However, unlike 1st and 2nd class, the survival proportion is significantly lower than 1.0, showing that many 3rd-class passengers did not survive.
    -   Crew: The female crew members have a much higher survival proportion than male crew members. This trend was already visible in q3, but now it’s clearer that the survival chance for male crew members was very low.
-   Comparison with q3:
    -   1st and 2nd class: In q3, female survivors outnumbered male survivors in absolute numbers. However, in this plot, the proportions are equal (1.0), which might be discrepancy.
    -   3rd class: In q3, the number of male and female survivors looked somewhat balanced, but this plot shows that the survival rate for males in 3rd class was much lower than for females. This suggests a much lower survival rate for males in this class, possibly due to restricted lifeboat access.
    -   Crew: In the q3 plot, the number of male survivors was significantly higher than the number of female survivors. However, in this plot, the survival proportion of female crew members is significantly higher than male crew members. This signifies that most female crew numbers who were less in number were able to survive unlike the male crew members which outnumbered the female crew members but most of them were not able to survive.
-   Is there anything *fishy* going on in your plot?
    -   Yes, there might be some discrepancies with the crew data.
    -   1st & 2nd Class Proportions Being 1.0: This suggests that all recorded individuals in these groups survived, which is historically inaccurate. There were non-survivors in both classes, so this could indicate a data filtering error or incorrect proportion calculations.

### **q5** Create a plot showing the group-proportion of occupants who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!

```{r q5-task}
ggplot(data = df_prop %>% filter(Survived == "Yes"), aes(x = Class, y = Prop, fill = Sex)) +
  geom_bar(stat = "identity", position = "dodge", na.rm = TRUE) +
  facet_grid(~Age) +
  labs(
    title = "Proportion of Survivors by Class, Sex, and Age",
    x = "Class",
    y = "Survival Proportion",
    fill = "Sex"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    strip.text = element_text(size = 12, face = "bold")
  )
```

**Observations**:

-   The bar plot represents the proportion of survivors, categorized by Class (1st, 2nd, 3rd, Crew) and further divided by Age group (Adult, Child) and Sex (Male, Female).
-   Observations for Adults:
    -   1st class adult females had the highest survival rate (\~95%), while 2nd class adult males had the lowest survival proportion (\~10%) out of all the categories.
    -   Females had a higher survival rate in all the classes when compared to the adult males who had significantly lower survival rates. This indicates that females were prioritized to get on the lifeboats first after the accident.
    -   1st-class adults had the best survival rate, likely due to better access to lifeboats.
    -   3rd-class adults had the lowest survival proportions, reflecting that lower-class passengers had limited access to lifeboats.
    -   Crew adult females had a significantly higher survival rate compared to the crew adult males, whose survival rate was quite low.
-   Observations for Children:
    -   Survival proportions for children in 1st and 2nd class were 100% for both male and female children, meaning every child in these classes survived and both these groups had the best survival rates out of all the proportion groups.
    -   3rd class children had significantly lower survival rates when compared to the other classes, and specifically, male children had lower survival rates than female children. This is probably because 3rd class children did not have an equitable access to the lifeboats when compared to 1st and 2nd class children.
-   If you saw something *fishy* in q4 above, use your new plot to explain the fishy-ness.
    -   I think the fishy-ness was caused because the proportions of adult survivors were being shadowed over by the data for the proportions of child survivors. So, for the 1st, 2nd, and 3rd classes, the data in the plot in q4 represented only the survival proportions for the children and not the adults. Since there was no children in the crew, the plot in q4 correctly represented the survival proportions for this class. Dividing the data by Age group (Adult, Child) enabled to resolve this fishy-ness because now we can see the data separately and the children data is not overlapping the adults data.

# Notes

<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.
