---
title: "Antibiotics"
author: "Sparsh Gupta"
date: 2025-03-03
output:
  github_document
---

*Purpose*: Creating effective data visualizations is an *iterative* process; very rarely will the first graph you make be the most effective. The most effective thing you can do to be successful in this iterative process is to *try multiple graphs* of the same data.

Furthermore, judging the effectiveness of a visual is completely dependent on *the question you are trying to answer*. A visual that is totally ineffective for one question may be perfect for answering a different question.

In this challenge, you will practice *iterating* on data visualization, and will anchor the *assessment* of your visuals using two different questions.

*Note*: Please complete your initial visual design **alone**. Work on both of your graphs alone, and save a version to your repo *before* coming together with your team. This way you can all bring a diversity of ideas to the table!

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|-------------------|----------------------------|-------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

```{r setup}
library(tidyverse)
library(ggrepel)
```

*Background*: The data[1] we study in this challenge report the [*minimum inhibitory concentration*](https://en.wikipedia.org/wiki/Minimum_inhibitory_concentration) (MIC) of three drugs for different bacteria. The smaller the MIC for a given drug and bacteria pair, the more practical the drug is for treating that particular bacteria. An MIC value of *at most* 0.1 is considered necessary for treating human patients.

These data report MIC values for three antibiotics---penicillin, streptomycin, and neomycin---on 16 bacteria. Bacteria are categorized into a genus based on a number of features, including their resistance to antibiotics.

```{r load}
## NOTE: If you extracted all challenges to the same location,
## you shouldn't have to change this filename
filename <- "./data/antibiotics.csv"

## Load the data
df_antibiotics <- read_csv(filename)
df_antibiotics %>% knitr::kable()
```

# Visualization

<!-- -------------------------------------------------- -->

### **q1** Prototype 5 visuals

To start, construct **5 qualitatively different visualizations of the data** `df_antibiotics`. These **cannot** be simple variations on the same graph; for instance, if two of your visuals could be made identical by calling `coord_flip()`, then these are *not* qualitatively different.

For all five of the visuals, you must show information on *all 16 bacteria*. For the first two visuals, you must *show all variables*.

*Hint 1*: Try working quickly on this part; come up with a bunch of ideas, and don't fixate on any one idea for too long. You will have a chance to refine later in this challenge.

*Hint 2*: The data `df_antibiotics` are in a *wide* format; it may be helpful to `pivot_longer()` the data to make certain visuals easier to construct.

```{r}
df_long <- df_antibiotics %>%
  pivot_longer(
    cols = c(penicillin, streptomycin, neomycin),
    names_to = "antibiotic", values_to = "MIC"
  )
```

#### Visual 1 (All variables)

In this visual you must show *all three* effectiveness values for *all 16 bacteria*. This means **it must be possible to identify each of the 16 bacteria by name.** You must also show whether or not each bacterium is Gram positive or negative.

```{r q1.1}
df_long %>%
  ggplot(aes(antibiotic, bacteria, fill = MIC)) +
  geom_tile() +
  scale_fill_viridis_c(trans = "log10") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  labs(
    title = "MIC values of Antibiotics for Bacteria",
    x = "Antibiotic", 
    y = "Bacteria"
  ) +
  facet_wrap(~gram, scales = "free_y")
```

#### Visual 2 (All variables)

In this visual you must show *all three* effectiveness values for *all 16 bacteria*. This means **it must be possible to identify each of the 16 bacteria by name.** You must also show whether or not each bacterium is Gram positive or negative.

Note that your visual must be *qualitatively different* from *all* of your other visuals.

```{r q1.2}
df_long %>%
  ggplot(aes(bacteria, MIC, fill = antibiotic)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_y_log10() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(
    title = "MIC Values of Antibiotics for Each Bacteria",
    x = "Bacteria", 
    y = "MIC (log scale)", 
    fill = "Antibiotic"
  ) +
  facet_wrap(~gram)
```

#### Visual 3 (Some variables)

In this visual you may show a *subset* of the variables (`penicillin`, `streptomycin`, `neomycin`, `gram`), but you must still show *all 16 bacteria*.

Note that your visual must be *qualitatively different* from *all* of your other visuals.

```{r q1.3}
df_long %>%
  ggplot(aes(antibiotic, MIC, fill = gram)) +
  geom_boxplot() +
  scale_y_log10() +
  labs(
    title = "Distribution of MIC Values by Antibiotic",
    x = "Antibiotic", 
    y = "MIC (log scale)", 
    fill = "Gram Staining"
  )
```

Inspecting the one extreme outlier for penicillin in visual 3 (box plot)

```{r}
penicillin_gp <- df_long %>% filter(antibiotic == "penicillin", gram == "positive")

Q1 <- quantile(penicillin_gp$MIC, 0.25)
Q3 <- quantile(penicillin_gp$MIC, 0.75)
IQR_value <- Q3 - Q1
upper_bound <- Q3 + 1.5 * IQR_value

outlier_bacteria <- penicillin_gp %>% filter(MIC > upper_bound)

print(outlier_bacteria)
```

#### Visual 4 (Some variables)

In this visual you may show a *subset* of the variables (`penicillin`, `streptomycin`, `neomycin`, `gram`), but you must still show *all 16 bacteria*.

Note that your visual must be *qualitatively different* from *all* of your other visuals.

```{r q1.4}
df_antibiotics %>%
  ggplot(aes(penicillin, streptomycin, color = gram)) +
  geom_point(size = 3) +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    title = "Penicillin vs Streptomycin MIC Values",
    x = "Penicillin MIC (log scale)", 
    y = "Streptomycin MIC (log scale)"
  )
```

#### Visual 5 (Some variables)

In this visual you may show a *subset* of the variables (`penicillin`, `streptomycin`, `neomycin`, `gram`), but you must still show *all 16 bacteria*.

Note that your visual must be *qualitatively different* from *all* of your other visuals.

```{r q1.5}
df_long %>%
  ggplot(aes(antibiotic, MIC, group = bacteria, color = gram)) +
  geom_line(aes(linetype = gram)) +
  geom_point() +
  scale_y_log10() +
  theme_minimal() +
  labs(
    title = "MIC Values Across Antibiotics",
    x = "Antibiotic", 
    y = "MIC (log scale)", 
    color = "Gram Staining"
  )
```

### **q2** Assess your visuals

There are **two questions** below; use your five visuals to help answer both Guiding Questions. Note that you must also identify which of your five visuals were most helpful in answering the questions.

*Hint 1*: It's possible that *none* of your visuals is effective in answering the questions below. You may need to revise one or more of your visuals to answer the questions below!

*Hint 2*: It's **highly unlikely** that the same visual is the most effective at helping answer both guiding questions. **Use this as an opportunity to think about why this is.**

#### Guiding Question 1

> How do the three antibiotics vary in their effectiveness against bacteria of different genera and Gram stain?

**Observations:**

What is your response to the question above?

-   The MIC values provide information about antibiotic effectiveness. Lower MIC values indicate higher effectiveness.

-   Penicillin appears to be more effective against Gram-positive bacteria (lower MIC values in most cases except for one outlier bacteria - *Streptococcus fecalis*), while neomycin and streptomycin show varied effectiveness across both Gram-positive and Gram-negative bacteria. We can also see that penicllin is least effective on Gram-negative bacteria when compared to neomycin and streptomycin.

-   Visual 1 (the heatmap), visual 2 (the bar plot), and the visual 5 (line chart) clearly show differences in MIC values across antibiotics and bacterial genera.

-   Visual 3 (the box plot) demonstrates the distribution of the effectiveness based on Gram staining and visually distinguishes antibiotic performance.

Which of your visuals above (1 through 5) is **most effective** at helping to answer this question?

-   Visual 3

Why?

-   Visual 3 effectively summarizes MIC distribution across Gram-positive and Gram-negative bacteria for each antibiotic.

-   It provides a direct comparison between the antibiotics, showing that penicillin generally has lower MIC values for Gram-positive bacteria while neomycin and streptomycin have more varied distributions.

-   Unlike the other plots, visual 3 provides a clearer overall trend, rather than individual values for each bacterium.

#### Guiding Question 2

In 1974 *Diplococcus pneumoniae* was renamed *Streptococcus pneumoniae*, and in 1984 *Streptococcus fecalis* was renamed *Enterococcus fecalis* [2].

> Why was *Diplococcus pneumoniae* was renamed *Streptococcus pneumoniae*?

**Observations:**

What is your response to the question above?

-   The renaming suggests that *Diplococcus pneumoniae* was reclassified into the streptococcus genus, likely due to genetic, structural, or functional similarities.

-   *Streptococcus pneumoniae* shares more characteristics with other *Streptococcus* species, possibly in morphology (chain-forming cocci rather than diplococci) or genetic sequencing.

Which of your visuals above (1 through 5) is **most effective** at helping to answer this question?

-   Visual 1

Why?

-   The heatmap groups bacteria into Gram-positive and Gram-negative categories.

-   We can see within the heatmap for the Gram-positive bacteria, the effect (MIC values) of the antibiotics on *Diplococcus pneumoniae* looks very similar to what the antibiotics have on two bacteria from the *Streptococcus* species, *Streptococcus viridans* and *Streptococcus hemolyticus*.

-   This suggests that *Diplococcus pneumoniae* shares antibiotic response characteristics with *Streptococcus*, reinforcing its reclassification.

-   While this plot or dataset is not enough information to obtain a direct taxonomic reason, they support the idea that both bacteria respond similarly to antibiotics, possibly due to shared cell wall structures or genetic similarities.

# References

<!-- -------------------------------------------------- -->

[1] Neomycin in skin infections: A new topical antibiotic with wide antibacterial range and rarely sensitizing. Scope. 1951;3(5):4-7.

[2] Wainer and Lysen, "That's Funny..." *American Scientist* (2009) [link](https://www.americanscientist.org/article/thats-funny)
