---
title: "Massachusetts Highway Stops"
author: "Sparsh Gupta"
date: 2025-04-11
output:
  github_document:
    toc: true
---

*Purpose*: In this last challenge we'll focus on using logistic regression to study a large, complicated dataset. Interpreting the results of a model can be challenging---both in terms of the statistics and the real-world reasoning---so we'll get some practice in this challenge.

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|------------------|-----------------------------|-------------------------|
| Effort | Some task **q**'s left unattempted | All task **q**'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission

<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**

*Background*: We'll study data from the [Stanford Open Policing Project](https://openpolicing.stanford.edu/data/), specifically their dataset on Massachusetts State Patrol police stops.

```{r setup}
library(tidyverse)
library(broom)
```

# Setup

<!-- -------------------------------------------------- -->

### **q1** Go to the [Stanford Open Policing Project](https://openpolicing.stanford.edu/data/) page and download the Massachusetts State Police records in `Rds` format. Move the data to your `data` folder and match the `filename` to load the data.

*Note*: An `Rds` file is an R-specific file format. The function `readRDS` will read these files.

```{r q1-task}
## TODO: Download the data, move to your data folder, and load it
filename <- "./data/yg821jf8611_ma_statewide_2020_04_01.rds"
df_data <- readRDS(filename)
```

# EDA

<!-- -------------------------------------------------- -->

### **q2** Do your "first checks" on the dataset. What are the basic facts about this dataset?

```{r}
summary(df_data)
head(df_data)
```

**Observations**:

-   Around 3.42 million total observations

<!-- -->

-   Data covers dates from January 1, 2007 to December 31, 2015

-   Subject age ranges from 10 to 94 years old, with a median age of 34

-   Most vehicles registered in Massachusetts (\~3.05 million), followed by CT, NY, NH, RI

-   Many variables have significant missing data, especially those related to contraband and searches

Note that we have both a `subject_race` and `race_Raw` column. There are a few possibilities as to what `race_Raw` represents:

-   `race_Raw` could be the race of the police officer in the stop
-   `race_Raw` could be an unprocessed version of `subject_race`

Let's try to distinguish between these two possibilities.

### **q3** Check the set of factor levels for `subject_race` and `raw_Race`. What do you note about overlap / difference between the two sets?

```{r q3-task}
## TODO: Determine the factor levels for subject_race and raw_Race
sort(unique(df_data$subject_race))
sort(unique(df_data$raw_Race))
```

**Observations**:

-   What are the unique values for `subject_race`?

    -   asian/pacific islander

    <!-- -->

    -   black

    -   hispanic

    -   white

    -   other

    -   unknown

-   What are the unique values for `raw_Race`?

    -   A

    <!-- -->

    -   American Indian or Alaskan Native

    -   Asian or Pacific Islander

    -   Black

    -   Hispanic

    -   Middle Eastern or East Indian (South Asian)

    -   None - for no operator present citations only

    -   White

-   What is the overlap between the two sets?

    -   Black

    <!-- -->

    -   Hispanic

    -   White

    -   Asian or Pacific Islander (maps to asian/pacific islander with standardization)

-   What is the difference between the two sets?

    -   `subject_race` uses standardized and grouped categories with consistent formatting

    <!-- -->

    -   `raw_Race` includes more granular or specific labels (e.g. "Middle Eastern or East Indian (South Asian)", "American Indian or Alaskan Native")

    -   `raw_Race` includes administrative categories like "None - for no operator present citations only"

    -   `raw_Race` has abbreviations like "A" that don’t appear in `subject_race`

### **q4** Check whether `subject_race` and `raw_Race` match for a large fraction of cases. Which of the two hypotheses above is most likely, based on your results?

*Note*: Just to be clear, I'm *not* asking you to do a *statistical* hypothesis test.

```{r q4-task}
## TODO: Devise your own way to test the hypothesis posed above.
df_data <- df_data %>%
  mutate(
    raw_Race_simplified = case_when(
      raw_Race %in% c("Asian or Pacific Islander") ~ "asian/pacific islander",
      raw_Race %in% c("Black") ~ "black",
      raw_Race %in% c("Hispanic") ~ "hispanic",
      raw_Race %in% c("White") ~ "white",
      raw_Race %in% c("American Indian or Alaskan Native", 
                      "Middle Eastern or East Indian (South Asian)") ~ "other",
      raw_Race %in% c("None - for no operator present citations only", "A") ~ "unknown",
      TRUE ~ NA_character_
    )
  )

df_data <- df_data %>%
  mutate(race_match = subject_race == raw_Race_simplified)

race_match_summary <- df_data %>%
  count(race_match) %>%
  mutate(percentage = n / sum(n) * 100)

print(race_match_summary)
```

**Observations**

Between the two hypotheses:

-   `race_Raw` could be the race of the police officer in the stop
-   `race_Raw` could be an unprocessed version of `subject_race`

which is most plausible, based on your results?

-   The most plausible explanation is that `raw_Race` is an unprocessed version of `subject_race`.

<!-- -->

-   Over 98% of the entries in the dataset show a match between `subject_race` and a simplified version of `raw_Race`. This strong agreement would not be expected if `raw_Race` referred to the race of the police officer, since we would then expect the values to differ from `subject_race` in most cases.

-   Additionally, the content of `raw_Race` includes fine-grained racial or ethnic categories that appear to be standardized or grouped in `subject_race`, further supporting the idea that `subject_race` is a cleaned or normalized version of `raw_Race`.

## Vis

<!-- ------------------------- -->

### **q5** Compare the *arrest rate*---the fraction of total cases in which the subject was arrested---across different factors. Create as many visuals (or tables) as you need, but make sure to check the trends across all of the `subject` variables. Answer the questions under *observations* below.

(Note: Create as many chunks and visuals as you need)

```{r}
df_data <- df_data %>%
  mutate(arrest_made = ifelse(is.na(arrest_made), FALSE, arrest_made))

df_data %>%
  filter(!is.na(subject_age)) %>%
  mutate(age_bin = cut(subject_age, breaks = seq(10, 90, by = 5))) %>%
  group_by(age_bin) %>%
  summarize(arrest_rate = mean(arrest_made)) %>%
  ggplot(aes(x = age_bin, y = arrest_rate)) +
  geom_col() +
  labs(title = "Arrest Rate by Subject Age", x = "Age Bin", y = "Arrest Rate") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

df_data %>%
  filter(!is.na(subject_sex)) %>%
  group_by(subject_sex) %>%
  summarize(arrest_rate = mean(arrest_made)) %>%
  ggplot(aes(x = subject_sex, y = arrest_rate, fill = subject_sex)) +
  geom_col() +
  labs(title = "Arrest Rate by Subject Sex", x = "Sex", y = "Arrest Rate") +
  theme(legend.position = "none")

df_data %>%
  filter(!is.na(subject_race)) %>%
  group_by(subject_race) %>%
  summarize(arrest_rate = mean(arrest_made)) %>%
  ggplot(aes(x = reorder(subject_race, -arrest_rate), y = arrest_rate, fill = subject_race)) +
  geom_col() +
  labs(title = "Arrest Rate by Subject Race", x = "Race", y = "Arrest Rate") +
  theme(legend.position = "none")
```

**Observations**:

-   How does `arrest_rate` tend to vary with `subject_age`?

    -   Arrest rates are highest among the youngest drivers (especially ages 10–15 and 25–35), then steadily decline with age.

    <!-- -->

    -   Drivers over 60 have much lower arrest rates.

-   How does `arrest_rate` tend to vary with `subject_sex`?

    -   Males have a significantly higher arrest rate than females.

-   How does `arrest_rate` tend to vary with `subject_race`?

    -   Hispanic and Black individuals have the highest arrest rates.

    <!-- -->

    -   White and Asian/Pacific Islander individuals have notably lower arrest rates.

# Modeling

<!-- -------------------------------------------------- -->

We're going to use a model to study the relationship between `subject` factors and arrest rate, but first we need to understand a bit more about *dummy variables*

### **q6** Run the following code and interpret the regression coefficients. Answer the the questions under *observations* below.

```{r q6-task}
## NOTE: No need to edit; inspect the estimated model terms.
fit_q6 <-
  glm(
    formula = arrest_made ~ subject_age + subject_race + subject_sex,
    data = df_data %>%
      filter(
        !is.na(arrest_made),
        subject_race %in% c("white", "black", "hispanic")
      ),
    family = "binomial"
  )

fit_q6 %>% tidy()
```

**Observations**:

-   Which `subject_race` levels are included in fitting the model?
    -   The model includes white, black, and hispanic individuals (others were filtered out).
-   Which `subject_race` levels have terms in the model?
    -   Only hispanic and white have coefficients; black is the reference category (baseline group), so it does not appear as a separate term.

You should find that each factor in the model has a level *missing* in its set of terms. This is because R represents factors against a *reference level*: The model treats one factor level as "default", and each factor model term represents a change from that "default" behavior. For instance, the model above treats `subject_sex==male` as the reference level, so the `subject_sexfemale` term represents the *change in probability* of arrest due to a person being female (rather than male).

The this reference level approach to coding factors is necessary for [technical reasons](https://www.andrew.cmu.edu/user/achoulde/94842/lectures/lecture10/lecture10-94842.html#why-is-one-of-the-levels-missing-in-the-regression), but it complicates interpreting the model results. For instance; if we want to compare two levels, neither of which are the reference level, we have to consider the difference in their model coefficients. But if we want to compare all levels against one "baseline" level, then we can relevel the data to facilitate this comparison.

By default `glm` uses the first factor level present as the reference level. Therefore we can use `mutate(factor = fct_relevel(factor, "desired_level"))` to set our `"desired_level"` as the reference factor.

### **q7** Re-fit the logistic regression from q6 setting `"white"` as the reference level for `subject_race`. Interpret the the model terms and answer the questions below.

```{r q7-task}
## TODO: Re-fit the logistic regression, but set "white" as the reference
## level for subject_race
library(forcats)

fit_q7 <- glm(
  formula = arrest_made ~ subject_age + subject_race + subject_sex,
  data = df_data %>%
    filter(
      !is.na(arrest_made),
      subject_race %in% c("white", "black", "hispanic")
    ) %>%
    mutate(subject_race = fct_relevel(subject_race, "white")),
  family = "binomial"
)


fit_q7 %>% tidy()
```

**Observations**:

-   Which `subject_race` level has the highest probability of being arrested, according to this model? Which has the lowest probability?
    -   Hispanic individuals have the highest probability of being arrested; white individuals have the lowest, according to the model.
-   What could explain this difference in probabilities of arrest across race? List **multiple** possibilities.
    -   Possible explanations include systemic bias in policing, differences in stop circumstances, differences in geographic or vehicle-related factors, or unobserved variables like behavior during the stop.
-   Look at the set of variables in the dataset; do any of the columns relate to a potential explanation you listed?
    -   The dataset includes columns like reason_for_stop, vehicle_type, search_conducted, and contraband_found, which could help explain differences in arrest rates across racial groups.

One way we can explain differential arrest rates is to include some measure indicating the presence of an arrestable offense. We'll do this in a particular way in the next task.

### **q8** Re-fit the model using a factor indicating the presence of contraband in the subject's vehicle. Answer the questions under *observations* below.

```{r q8-task}
## TODO: Repeat the modeling above, but control for whether contraband was found
## during the police stop
fit_q8 <- glm(
  formula = arrest_made ~ subject_age + subject_race + subject_sex + contraband_found,
  data = df_data %>%
    filter(
      !is.na(arrest_made),
      subject_race %in% c("white", "black", "hispanic")
    ) %>%
    mutate(
      subject_race = fct_relevel(subject_race, "white"),
      contraband_found = ifelse(is.na(contraband_found), FALSE, contraband_found)
    ),
  family = "binomial"
)

fit_q8 %>% tidy()
```

**Observations**:

-   How does controlling for found contraband affect the `subject_race` terms in the model?
    -   Controlling for contraband slightly reduces the coefficients for both black and hispanic subjects, but substantial racial disparities in arrest probability remain.
-   What does the *finding of contraband* tell us about the stop? What does it *not* tell us about the stop?
    -   Finding contraband indicates that an arrestable offense may have occurred, which helps explain some arrests.
    -   It does not tell us whether the search itself was justified or conducted fairly, nor does it explain why some groups are searched more often than others.

### **q9** Go deeper: Pose at least one more question about the data and fit at least one more model in support of answering that question.

**Question:** Are racial disparities in arrests primarily driven by differences in search rates rather than differences in actual behavior or contraband possession?

```{r q9}
fit_q9 <- glm(
  formula = search_conducted ~ subject_age + subject_race + subject_sex,
  data = df_data %>%
    filter(
      !is.na(search_conducted),
      subject_race %in% c("white", "black", "hispanic")
    ) %>%
    mutate(subject_race = fct_relevel(subject_race, "white")),
  family = "binomial"
)

fit_q9 %>% tidy()
```

**Observations**:

-   Black and Hispanic individuals are significantly more likely to be searched than white individuals, even after controlling for age and sex.
-   Hispanic subjects have the highest predicted likelihood of being searched, followed by Black subjects.

<!-- -->

-   Younger people are more likely to be searched than older individuals.

-   Females are less likely to be searched than males.

-   These disparities in search rates may contribute to higher arrest rates among certain racial groups, even if contraband possession is similar.

## Further Reading

<!-- -------------------------------------------------- -->

-   Stanford Open Policing Project [findings](https://openpolicing.stanford.edu/findings/).
